---
title: "MEL differential expression analysis and more"
author: "[Aristeidis Sionakidis](https://www.ed.ac.uk/profile/aristeidis-sionakidis), Aliki Papadimitriou"
date: "`r Sys.Date()`"
abstract: >
  This markdown document contains all the code from the R scripts of this
  repository. The code may be slightly modified so that the document has a
  smooth, natural flow for the reader, but all results generated are identical.
  The purpose of this markdown is to visualize the results, explain the
  rationale behind all the preprocessing steps and main analysis steps we take,
  and make it easy to go through our work and understand our approach. In the end
  of the document we provide our `sessionInfo()` output for reproducibility.
  This work focuses on the analysis of RNA-seq count data from MEL samples
  which either underwent treatment with HMBA or not (controls). Samples are
  sequenced in multiple timepoints. There are two batches (`2023_exp`, `auth`)
  for which results are adjusted.
output: 
  html_document:
    fig_width: 10
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, dev = "png", tidy = FALSE,
                      message = FALSE, error = FALSE, warning = TRUE)
```

# Libraries
```{r libraries}
# Loading all libraries required for the workflow
suppressWarnings(library(dplyr))
suppressWarnings(library(DESeq2))
suppressWarnings(library(edgeR))
suppressWarnings(library(org.Mm.eg.db))
suppressWarnings(library(AnnotationHub))
suppressWarnings(library(ggplot2))
suppressWarnings(library(tidyr))
suppressWarnings(library(vsn))
suppressWarnings(library(ggpubr))
suppressWarnings(library(cowplot))
suppressWarnings(library(grid))
suppressWarnings(library(pheatmap))
suppressWarnings(library(rcartocolor))
suppressWarnings(library(RColorBrewer))
suppressWarnings(library(tibble))
suppressWarnings(library(PoiClaClu))
suppressWarnings(library(ggrepel))
suppressWarnings(library(glmpca))
suppressWarnings(library(openxlsx))
suppressWarnings(library(EnhancedVolcano))
suppressWarnings(library(genefilter))
suppressWarnings(library(Matrix))
```

# Preprocessing: First steps
## Loading and cleaning files
All the files are in the subdirectory `Count_files/`. We first load all of their names and then read them all and combine them in a single object using the function`readDGE()` from the `edgeR` package.
```{r loading}
# Load the files into R
files = list.files("Count_files/")
x = readDGE(paste0("Count_files/", files), 
            columns = c(1,2))
dim(x)
```
Our object contains a count expression matrix with 24983 features (genes) and 14 samples. We first check the sample names, which are currently very long (full directory locations).
```{r print_colnames_pre}
# Print current column names of x (sample names)
colnames(x)
```
We shorten the names to make them more readable and meaningful.

```{r sample names}
sample_names_prev = colnames(x)
samplenames_almost = substring(sample_names_prev, nchar("Count_files/") + 1,
                               nchar(sample_names_prev) - 7)

# Add prefix to indicate batch
samplenames = samplenames_almost
for (i in 1:ncol(x)) {
  if (i <= 8) {
    samplenames[i] = paste0("2023_", samplenames[i])
  } else {
    samplenames[i] = paste0("AUTH_", samplenames[i])
  }
}

colnames(x) = samplenames
rm(sample_names_prev, samplenames_almost, i, samplenames)
```
The updated sample names look better:
```{r updated_sample_names}
colnames(x)
```
As mentioned in the abstract, the data come from two separate batches (`2023_exp` and `auth`). We add this information in our objects phenotypic data. We also add information about the treatment and the timepoint of sample collection.

```{r batch}
# Create batch variable
batch = c(rep("2023_exp", 8), rep("auth", 6)) # 2 batches

HMBA_ind = grepl("HMBA", colnames(x))
treatment = rep("control", 14)
treatment[HMBA_ind] = "HMBA" # 2 treatments

timepoint = c("24h", "24h", "48h", "48h", "72h", "72h", "48h", "48h",
              "24h", "48h", "72h", "24h", "48h", "72h")

# Add to the DGE object
x$samples$batch = factor(batch)
x$samples$batch = relevel(x$samples$batch, ref = "auth")
x$samples$treatment = factor(treatment)
x$samples$treatment = relevel(x$samples$treatment, ref = "control")
x$samples$timepoint = factor(timepoint)
x$samples$timepoint = relevel(x$samples$timepoint, ref = "24h")
rm(batch, treatment, timepoint, HMBA_ind)
```

## Additional gene annotation
We already have Gene Symbol annotation for the rows of each sample's library. However, we will add NCBI Entrez ID's as well so that at least one additional annotation is available. The symbols to Entrez relationship is not a 1:1 relationship, however we use here the `org.Mm.eg.db` to do the mapping of symbols to their official Entrez identifier, wherever this is possible.

```{r entrez}
# Map the genes to Entrez IDs so that we have at least two annotations available
library(org.Mm.eg.db)
official = org.Mm.egSYMBOL
mapped_genes_official = mappedkeys(official)
official_df = as.data.frame(official[mapped_genes_official])
official_df = official_df %>% dplyr::rename(EntrezGene.ID = gene_id, Gene.Symbol = symbol)
official_df$HGNC_Official = "Yes"
official_df = official_df[-which(duplicated(official_df$Gene.Symbol)), ]

rm(official, mapped_genes_official, files)

# Create a genes slot with gene annotation
gene_annot = as.data.frame(list(Gene.Symbol = rownames(x))) %>%
  left_join(official_df, by = "Gene.Symbol") %>%
  dplyr::select(-HGNC_Official)
x$genes = gene_annot
```

## Data filtering
It's common in RNA-seq experiments to find row in a count matrix which are completely filled with zero's. These rows are uninformative and only add noise to downstream analytical steps.

```{r zeros}
# How many genes in this dataset are all zeros?
table(rowSums(x$counts==0)==14) # 14: number of samples
```

```{r print_zero_percentage}
print(paste0(round(100*8463/(16520+8463), 2), "% of genes only have zero counts."))
```
A considerable number of rows (8463, ~34%) must be removed from the matrix. Along with these rows, we must remove other noisy rows, e.g. rows which are mostly zero, or rows with only zeros on particular subgroups of interest for our analysis. In order to do that we use the `filterbyExpr()` function of `edgeR`.

`filterByExpr()` is used to filter out genes (or other countable features like transcripts) that are too lowly expressed to provide reliable statistical inference in the context of a differential expression analysis.

The function calculates a threshold for the minimum count that a gene should have across the samples in the experiment. The threshold is calculated based on the library sizes and the complexity of the experimental design. Genes that don't meet this threshold are filtered out.

The function uses the following criteria by default:

  1. Each gene must have a count-per-million (CPM) above a certain threshold (by default, log~2~(CPM) of 0.5) in a certain number of samples. For more information on CPM, RPKM, FPKM and TPM check [here](https://www.rna-seqblog.com/rpkm-fpkm-and-tpm-clearly-explained/).
  2. The "certain number of samples" is determined by the experimental design. For a simple two-group comparison, it would be the size of the smaller group. For a more complex design, it would be the smallest number of samples that can give a degree of freedom in the linear model fit. In other words, the smallest group of samples that can possibly show differential expression.

In our case, we are using a complex experimental design with interaction terms. The function will look at this design and determine the smallest number of samples that can possibly show differential expression given this design. It will then filter out genes that don't have a sufficient count in at least this many samples. Let's firstly define the design matrix:

```{r design}
# Create a full design matrix for the filtering
design = model.matrix(~ x$samples$batch + x$samples$treatment + x$samples$timepoint +
                        x$samples$treatment:x$samples$timepoint)
```
Our design matrix contains an interaction term because we are interested in making nested comparisons later, based on results from differential expression analysis. The function then checks two criteria for each gene:

  1. CPM cutoff: The function calculates the CPM for each gene in each sample, and checks that each gene has a CPM above a certain <u>cutoff</u> in at least `MinSampleSize` (source code variable) samples. In our case, based on the source code, `MinSampleSize = 1`. In order to estimate `MinSampleSize`, the Hat matrix, $H$, of the design matrix $X$ must be computed as $$H = X(X^TX)^{-1}X^T$$ Here, $X^T$ denotes the transpose of the matrix $X$, and $(X^TX)^{-1}$ denotes the inverse of the matrix product $X^TX$. `hat()` computes $H$ from $X$ and outputs the diagonal entries of $H$, which measure the influence that each individual observation has on its own predicted value. Then, `MinSampleSize` is calculated as the ratio of 1 over the maximum value in the diagonal of $H$.
```{r hat}
# No need to run this code, it is automatically done by filterByExpr(). Only used for context here.
h = stats::hat(design)
h
```
```{r minsamplesize}
MinSampleSize = 1/max(hat(design))
MinSampleSize
```
  2. The <u>cutoff</u> is calculated as `min.count / MedianLibSize * 1e6`, where `min.count` is a minimum count (default is `10`), and `MedianLibSize` is the median library size (total count) across all samples.
  
  3. <u>Total count cutoff</u>: The function checks that the total count for each gene (summed across all samples) is above a certain cutoff (min.total.count, default is `15`).
  
We are using default values here.
```{r filterbyexpr}
# Remove h and MinSampleSize from the environment if you ran the previous two chunks
if (exists("h")) { rm(h) }
if (exists("MinSampleSize")) { rm(MinSampleSize) }

# Filter using fitlerByExprs()
keep.exprs = names(which(filterByExpr(x, design = design) == TRUE))
x_filt = x
x_filt$counts = x_filt$counts[keep.exprs,]
x_filt$genes = x_filt$genes[x_filt$genes$Gene.Symbol %in% keep.exprs,]
dim(x_filt)
```
We now have 12027 genes (out of the original 24983) that fulfill the default criteria of `filterByExprs()`.

## Generate a TPM (Transcripts Per Million) matrix
The transformation from counts to TPM requires the length of each gene to be known. In order to transform a count matrix to a TPM matrix we follow these three steps (taken again from [here](https://www.rna-seqblog.com/rpkm-fpkm-and-tpm-clearly-explained/):

  1. Divide the read counts by the length of each gene in kilobases (kb). This gives us reads per kilobase `rpk`.
  2. Count up all the `rpk` values in a sample and divide this number by 1,000,000. This is our “per million” scaling factor.
  3. Divide the `rpk` values by the “per million” scaling factor. This gives us TPM.
  
The sum of all TPMs in each sample is the same. This makes it easier to compare the <u>proportion</u> of reads that mapped to a gene in each sample. If the TPM for gene A in Sample 1 is 3.33 and the TPM in sample B is 3.33, then we know that the exact same proportion of total reads mapped to gene A in both samples. **That does not mean that the original counts for gene A were equal, but their proportion over all the counts for both samples, is the same.**

We use the `AnnotationHub` library (which automatically requires the `ensembldb` library too) in order to retrieve **transcript** (not full gene) lengths for the genes in our data. It is important to notice that there may be more than one length for a gene, transcript, coding sequence etc. and lengths may also change from time to time based on new knowledge. Here, for reproducibility, we keep the longest transcript length for each mapped transcript. After computing the TPM matrix, we also generate a log~2~-scaled version using the transformation `log2(tpm+1)` (the pseudocount of `1` here is arbitrary) which <u>sometimes</u> ([but not always](https://www.biorxiv.org/content/10.1101/2022.06.11.495771v2.full.pdf), check refs 4,18,34,35,37 if interested) produces a normal distribution.

```{r tpm}
# Create an AnnotationHub object
ah = AnnotationHub()

# Query for the latest Mus musculus EnsDb object
latest_mus_musculus = query(ah, c("EnsDb", "Mus musculus")) 

# Print the latest version
tail(latest_mus_musculus)

# ID of interest is "AH109650"
edb = ah[["AH109650"]]

# Get gene data
genes_data = genes(edb)
genes_map = data.frame(gene_id = genes_data$gene_id,
                       Gene.Symbol = genes_data$gene_name)

# Get transcript data
transcripts_data = transcripts(edb)
transcript_lengths = data.frame(transcript_id = transcripts_data$tx_id,
                                gene_id = transcripts_data$gene_id,
                                tx_length = width(transcripts_data)) %>%
  inner_join(genes_map, by = "gene_id") %>%
  dplyr::filter(!Gene.Symbol == "") %>%
  group_by(Gene.Symbol) %>%
  summarise(longest_transcript_length = max(tx_length, na.rm = TRUE))

# Calculate gene lengths in kilobases
transcript_lengths$transcript_length_kb = transcript_lengths$longest_transcript_length / 1000

# Map gene lengths to rownames of the counts matrix
rownames(transcript_lengths) = transcript_lengths$Gene.Symbol
mapped = transcript_lengths[transcript_lengths$Gene.Symbol %in% 
                                      x_filt$genes$Gene.Symbol,]
mapped_lengths = mapped$transcript_length_kb

# Calculate counts per kilobase (CPK)
cpk = x_filt$counts[mapped$Gene.Symbol, ] / mapped_lengths

# Calculate the sum of CPK values for each sample
cpk_sum = colSums(cpk)

# Calculate TPM values
tpm = cpk / cpk_sum * 1e6
log2tpm = log2(tpm + 1)
rm(cpk, cpk_sum, keep.exprs, mapped_lengths, ah, edb)
```

In our experiment there are two features which were included in the assay used, which are of no use, as one can tell by their names: `"no_feature", "ambiguous"`. We remove those.

```{r removals}
removals = c(which(x_filt$genes$Gene.Symbol == "no_feature"),
             which(x_filt$genes$Gene.Symbol == "ambiguous"))
x_filt$genes = x_filt$genes[-removals, ]
x_filt$counts = x_filt$counts[-removals, ]
```

That concludes the preprocessing. We now proceed with creating a `DESeq2` object for our analysis and the production of exploratory plots.

# Creating a DESeq2 object
We will use our `x_filt` object's data to generate a `DESeq2` object which will be handy for all types of analysis from now on.
```{r deseq2}
# DESeq2 object
dds = DESeqDataSetFromMatrix(countData = x_filt$counts,
                             colData = x_filt$samples,
                             design = ~ batch + treatment + timepoint +
                               treatment:timepoint)

featureData = as.data.frame(x_filt$genes)
mcols(dds) = DataFrame(mcols(dds), featureData)
mcols(dds)
rm(featureData)
```

# Exploratory analysis
## Distributions of data
**Note:** All plots generated here are not automatically exported as figures in the repository. However, the scripts in this repository provide additional code to produce high quality figures and export them.

In this section, we produce two types of plots for two different types of expression matrices (counts and `log2(tpm+1)`). Both types of plots display the distribution of expression values across all samples. In the first type (density curves), the overall distribution is plotted, while in the second type (color density plots), the distribution of each sample's values are overlaid on top of the others.

We define two functions which make use of `ggplot2` and `tidyr` functionality to generate our distribution plots and keep our code succinct.

```{r distribution_plot_functions}
# Density curve function
create_density_curve <- function(matrix) {
  as.data.frame(matrix) %>%
    gather(.) %>%
    ggplot(aes(x = value, fill = "Density Curve")) + 
    geom_density(alpha = 0.5, position = "identity") +
    xlim(-1, NA) +
    labs(title = "Distribution of values by sample",
         x = "Value", y = "Density") +
    theme_classic()+
    theme(legend.position = "none")
}

# Per column density plot
create_density_plot_color <- function(matrix) {
  as.data.frame(matrix) %>%
    gather(., key = "sample", value = "value") %>%
    ggplot(aes(x = value, fill = sample)) + 
    geom_density(alpha = 0.5, position = "identity") +
    xlim(-1, NA) +
    labs(title = "Distribution of values by sample",
         x = "Value", y = "Density") +
    theme_classic()+
    theme(legend.position = "none")
}
```

From now on, conditional statements such as the following will precede plot generation and exporting code. Their purpose is to create output directories if they do not already exist (e.g. in the case someone runs the code without downloading the project from [GitHub](https://github.com/sionaris/mel_auth)).

```{r first_cond_for_output_dir}
# Create folder for distribution plots
if (!dir.exists("MEL_exploratory_plots")) {
  dir.create("MEL_exploratory_plots")
}

if (!dir.exists("MEL_exploratory_plots/MEL_distribution_plots")) {
  dir.create("MEL_exploratory_plots/MEL_distribution_plots")
}
```

```{r count_densities_no_xlim}
# Density curve and sample densities for raw counts
ggarrange(create_density_curve(x_filt$counts), 
          create_density_plot_color(x_filt$counts), 
          ncol = 2, nrow = 1,
          labels = NULL,
          font.label = list(size = 4, color = "black", face = "bold"))
```

The previous plot contains count outliers (abnormally high counts) which make it difficult to see the distribution of counts near zero. We replot with `xlim` set to `200`.

```{r count_densities_with_xlim}
# Density curve and sample densities for raw counts
ggarrange(create_density_curve(x_filt$counts)+
  scale_x_continuous(limits = c(-1, 200), breaks = seq(0, 200, 20)), 
          create_density_plot_color(x_filt$counts)+
  scale_x_continuous(limits = c(-1, 200), breaks = seq(0, 200, 20)), 
          ncol = 2, nrow = 1,
          labels = NULL,
          font.label = list(size = 4, color = "black", face = "bold"))
```

Now we produce the same figures for the `log2(tpm+1)` data:
```{r tpm_densities}
# Density curve and sample densities for raw counts
ggarrange(create_density_curve(log2tpm), 
          create_density_plot_color(log2tpm), 
          ncol = 2, nrow = 1,
          labels = NULL,
          font.label = list(size = 4, color = "black", face = "bold"))
```

## Count data transformations
Two common count data transformations in the `DESeq2` package are the variance-stabilizing transformation (`vst`) and the regularized logarithm transformation (`rlog`). The point of these two transformations is to remove the <u>dependence of the variance on the mean</u>, particularly the high variance of the logarithm of count data when the mean is low. Both VST and rlog use the experiment-wide trend of variance over mean, in order to transform the data to remove the experiment-wide trend. Note that it is neither required nor desired that all the genes have exactly the same variance after transformation.

The VST is a technique to stabilize the variance across the range of count data in RNA-seq analysis. It is based on the properties of the negative binomial distribution which is often used to model RNA-seq count data. When `fitType="parametric"` is used, the dispersion-mean relationship is modeled as a parametric curve. This transformation aims to stabilize the variance across the range of count data. It helps to normalize the data in such a way that for large counts, the transformation becomes asymptotically equal to the logarithm (base 2) of normalized counts. The full mathematical process for the default case (`fitType="parametric"`) [here](https://github.com/mikelove/DESeq2/blob/devel/inst/script/vst.pdf).

The function `rlog()`, stands for regularized log, transforms the original count data to the log2 scale by fitting a model with a term for each sample and a prior distribution on the coefficients which is estimated from the data. The resulting data contains elements defined as: $$\log_2(q_{ij}) = \beta_{i0} + \beta_{ij}$$ where $q_{ij}$ is a parameter proportional to the expected true concentration of fragments for gene $i$ and sample $j$, $β_{i0}$ is an intercept which does not undergo shrinkage, and $β_{ij}$ is the sample-specific effect which is shrunk toward zero based on the dispersion-mean trend over the entire dataset. The trend typically captures high dispersions for low counts, and therefore these genes exhibit higher shrinkage from the `rlog()`. In conclusion, `rlog()` works in a way which minimizes differences between samples for rows with small counts, and which normalizes with respect to library size. The rlog transformation produces a similar variance stabilizing effect as `vst()`, though `rlog()` is more robust in the case when the size factors vary widely. However, it is slower than `vst()`, which may be important for big sample sizes.

```{r vst_rlog}
# VST and rlog transformations
vsd = vst(dds, blind = FALSE)
rld = rlog(dds, blind = FALSE)
```
See [here](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#data-transformations-and-visualization) for why we set `blind = FALSE`.

We now proceed by estimating size factors and producing mean - sd ($\sqrt{var}$) plots for three different transformations: $\log_2(counts + 1)$, $\text{vst}(counts)$ and $\text{rlog}(counts)$.

```{r mean_sd_plots}
dds = estimateSizeFactors(dds)

# Create folder for mean-variance plots and exploratory heatmaps
if (!dir.exists("MEL_exploratory_plots/Transformations_and_heatmaps")) {
  dir.create("MEL_exploratory_plots/Transformations_and_heatmaps")
}

# Produce a scatter plot of transformed counts for the first two samples
df = bind_rows(
  as.data.frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
    mutate(transformation = "log2(x + 1)"),
  as.data.frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as.data.frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))

colnames(df)[1:2] <- c("x", "y")  

lvls = c("log2(x + 1)", "vst", "rlog")
df$transformation = factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)

# Produce mean - sd plots 
ntd = normTransform(dds) # log2(counts + 1) transformation
plot1 = meanSdPlot(assay(ntd))$gg + rremove("ylab") + rremove("xlab") + 
  ggtitle(bquote(bold("log2(counts + 1)"))) +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.5))
plot2 = meanSdPlot(assay(rld))$gg + rremove("ylab") + rremove("xlab") + 
  ggtitle(bquote(bold("rlog"))) +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.5))
plot3 = meanSdPlot(assay(vsd))$gg + rremove("ylab") + rremove("xlab") + 
  ggtitle(bquote(bold("vst"))) +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.5))

# Use ggarrange to align the three plots horizontally and set the x-axis and y-axis labels
fig = ggarrange(plot1, plot2, plot3, 
                ncol = 3, nrow = 1, 
                common.legend = TRUE, legend = "bottom",
                labels = c("A", "B", "C"),
                align = "hv",
                font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))
annotate_figure(fig, left = textGrob(bquote(bold("sd")), rot = 90, vjust = 1, gp = gpar(cex = 1)),
                bottom = textGrob(bquote(bold("rank (mean)")), gp = gpar(cex = 1)))
```
The `rlog()` and `vst()` are shrinking the effect of noisy genes with low counts on the variance, more effectively than the `log2(counts+1)` transformation.

## Exploratory heatmaps
We now proceed with producing exploratory heatmaps using the three transformations we calculated in the previous steps. The heatmaps will contain information from the top 50 genes of each transformation. The top 50 genes are determined here based on the `rowMeans()` of the original counts. The `rowMedians()` can be used as well.

```{r heatmaps_expl_top50}
select = order(rowMeans(counts(dds,normalized=TRUE)),
               decreasing=TRUE)[1:50]
df = as.data.frame(colData(dds)[,c("batch", "treatment", "timepoint")])
```

We prepare custom color annotation for our variables:
```{r ann_colors}
# Annotation colors
ann_colors = list(
  batch = c(`2023_exp` = "dodgerblue4", auth = "deeppink4"),
  treatment = c(HMBA = "purple4", control = "orange"),
  timepoint = c(`24h` = carto_pal(n = 7, "Emrld")[1], 
                `48h` = carto_pal(n = 7, "Emrld")[4], 
                `72h` = carto_pal(n = 7, "Emrld")[7])
)

# log2(counts + 1)
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df, main = "log2(counts + 1)",
         annotation_colors = ann_colors, fontsize = 6)

# rlog
pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df, main = "rlog",
         annotation_colors = ann_colors, fontsize = 6)

# vst
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df, main = "vst",
         annotation_colors = ann_colors, fontsize = 6)
```

For all subsequent visualizations where transformed counts are used, we use the `vsd` object, i.e. the counts that under went VST. Next up, we visualize high-dimensional distances between samples based on the raw counts and VST data. For the VST data we use Euclidean distances (more suitable for continuous data such as VST data), while for the count matrix we use Poisson distance: this measure of dissimilarity between counts also takes the inherent variance structure of counts into consideration when calculating the distances between samples.

Here is the formula for the Poisson distance: $$D_{ij} = \sum_{k=1}^{n} \left( n_{0_{ik}} + n_{0_{jk}} - n_{0_{ik}} \times \frac{y_{ik} + \beta}{n_{0_{ik}} + \beta} - n_{0_{jk}} \times \frac{y_{jk} + \beta}{n_{0_{jk}} + \beta} + y_{ik} \times \log \left( \frac{y_{ik} + \beta}{n_{0_{ik}} + \beta} \right) + y_{jk} \times \log \left( \frac{y_{jk} + \beta}{n_{0_{jk}} + \beta} \right) \right)$$ where $D_{ij}$ is the Poisson distance between row $i$ and row $j$ of the transposed count matrix (samples $i$ and $j$), $n_{0_{ik}}$ and $n_{0_{jk}}$ are elements of the null model matrix $n0$, for rows $i$ and $j$, respectively, $y_{ik}$ and $y_{jk}$ are elements of the transformed count matrix $x$ (original count matrix $Y_{m\times n}$ raised to the power of alpha; alpha is chosen automatically based on maximum likelihood), for rows $i$ and $j$, respectively, $\beta$ is a smoothing parameter and $k$ runs from 1 to $n$, where $n$ is the number of columns in the count matrix.
```{r dist_heatmaps_1, fig.height=10}
# Euclidean distances for VST data
sampleVSTdists = dist(t(assay(vsd)))
sampleDistMatrix = as.matrix(sampleVSTdists)
colors = colorRampPalette(viridisLite::magma(10))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleVSTdists,
         clustering_distance_cols = sampleVSTdists,
         col = colors,
         fontsize = 8,
         main = "Sample distance heatmap",
         annotation_col = df,
         annotation_colors = ann_colors)

# Poisson distances for counts
poisd = PoissonDistance(t(counts(dds)))
samplePoisDistMatrix = as.matrix(poisd$dd)
dimnames(samplePoisDistMatrix) = dimnames(sampleDistMatrix)

pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors,
         fontsize = 8,
         main = "Sample Poisson distance heatmap",
         annotation_col = df,
         annotation_colors = ann_colors)
```
In both distance heatmaps it is evident that samples that underwent treatment are more similar to each other (small high-dimensional distances between them).

## Dimensionality reduction plots
In this section we produce principal component analysis (PCA) and multidimensional scaling (MDS) plots, in order to plot high-dimensional relationships between features and samples in 2D space. 

PCA identifies the directions (called principal components) in which the data varies the most, and projects the original data onto these new orthogonal axes. The first few principal components often capture much of the variation in the gene expression matrix, allowing us to visualize complex data in a simplified, lower-dimensional format. 

For a gene expression matrix $Y_{m\times n}$, PCA produces $m$ orthogonal axes which capture unique parts of the original variation in $Y$. If summed all together, they capture ~100% of the variation. Just because of the underlying methodology, the first principal component captures the highest proportion of original variation across components, the second component captures the second highest proportion etc. Therefore, we set our x axis to be the first principal component (PC1) and the y axis to be the second principal component (PC2). The coordinates of samples in each component (each axis) are determined by a separate linear combination of the $m$ gene variables. 

A <u>key limitation of PCA</u> is that it assumes the high-dimensional data are normally distributed and is therefore more appropriate for data such as VST or rlog-transformed matrices.

```{r pca}
# Create dimensionality reduction plots folder 
if (!dir.exists("MEL_exploratory_plots/Dimensionality_reduction_plots")) {
  dir.create("MEL_exploratory_plots/Dimensionality_reduction_plots")
}

# PCA on the VST data
pcaData = plotPCA(vsd, intgroup = c("treatment", "batch"), 
                  returnData=TRUE) %>%
  dplyr::rename(Sample.ID = name)
time = as.data.frame(vsd@colData@listData) %>%
  rownames_to_column %>%
  dplyr::select(Sample.ID = rowname, timepoint)
pcaData = pcaData %>%
  inner_join(time, by = "Sample.ID")
percentVar = round(100 * attr(pcaData, "percentVar"))
pcaplot = ggplot(pcaData, aes(PC1, PC2, color = treatment, shape = batch)) +
  geom_point(size = 2, alpha = 0.65) +
  geom_text_repel(aes(label = timepoint), segment.linetype = 1,
                      segment.color = "black", segment.size = 0.6,
                  min.segment.length = 0,
                  size = 3.5, nudge_y = 0.03, nudge_x = 0.03, show.legend = FALSE) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  ggtitle("Principal Component Analysis: VST") +
  theme_classic() +
  scale_color_manual(name = "treatment", values = c("orange", "purple4"),
                     labels = c("control", "HMBA"))+
  theme(plot.title = element_text(size = 12, face = "bold", vjust = 0.5, hjust = 0.5),
        axis.text = element_text(size = 8, hjust = 0.5, vjust = 0.5, 
                                 color = "black"),
        axis.title = element_text(size = 10, face = "bold"),
        axis.ticks = element_line(linewidth = 0.1),
        axis.line = element_line(linewidth = 0.3),
        legend.position = "right",
        legend.key.size = unit(6, units = "mm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face = "bold", size = 9),
        legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "mm"),
        legend.spacing.y = unit(2, units = "mm"),
        legend.spacing.x = unit(2, units = "mm"),
        legend.background = element_blank())
```

Another technique for performing dimension reduction on data that is <u>not normally distributed</u> (e.g. over-dispersed count data) is generalized principal component analysis, or GLM-PCA, as implemented in the CRAN package `glmpca`. The function' uses 's algorithm is based on the expectation-maximization (EM) algorithm, an iterative algorithm which depends on the initial values (which are randomly selected every time). To ensure reproducibility of results and the generation of identical plots, we use `set.seed(123)`. This ensures that randomizations will be the same across computers. To further ensure reproducibilty, we specify the random number generator version to `"4.2.2"` with `RNGversion()`.

```{r glmpca}
# GLM-PCA on raw counts
RNGversion("4.2.2")
set.seed(123)
gpca = glmpca(counts(dds), L=2)
gpca.dat = gpca$factors
gpca.dat$treatment = dds$treatment
gpca.dat$batch = dds$batch
gpca.dat$timepoint = dds$timepoint

gpcaplot = ggplot(gpca.dat, aes(x = dim1, y = dim2, color = treatment, shape = batch)) +
  geom_point(size = 2, alpha = 0.65) +
  geom_text_repel(aes(label = timepoint), segment.linetype = 1,
                      segment.color = "black", segment.size = 0.6,
                  min.segment.length = 0,
                  size = 3.5, nudge_y = 0.03, nudge_x = 0.03, show.legend = FALSE) +
  ggtitle("Generalized Principal Component Analysis: counts") +
  theme_classic() +
  scale_color_manual(name = "treatment", values = c("orange", "purple4"),
                     labels = c("control", "HMBA"))+
  theme(plot.title = element_text(size = 12, face = "bold", vjust = 0.5, hjust = 0.5),
        axis.text = element_text(size = 8, hjust = 0.5, vjust = 0.5, 
                                 color = "black"),
        axis.title = element_text(size = 10, face = "bold"),
        axis.ticks = element_line(linewidth = 0.1),
        axis.line = element_line(linewidth = 0.3),
        legend.position = "right",
        legend.key.size = unit(6, units = "mm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face = "bold", size = 9),
        legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "mm"),
        legend.spacing.y = unit(2, units = "mm"),
        legend.spacing.x = unit(2, units = "mm"),
        legend.background = element_blank())

ggarrange(pcaplot, gpcaplot, 
          ncol = 2, nrow = 1, common.legend = TRUE, 
          labels = c("A", "B"), legend = "right",
          font.label = list(size = 10, color = "black", face = "bold"))
```
In the left plot (PCA on the VST data), we see that the principal direction of variation is dominated by the treatment effect, as purple points (HMBA-treated samples) cluster to the left and yellow points (control samples) cluster to the right. There's also a clear separation in PC2, which is dominated by the batch effect, with samples from `batch = "auth"` at the top and samples from `batch = "2023_exp"` at the bottom. The effect of timepoint is mild, however samples from similar timepoints (for a given batch) seem to cluster together.

In the GPCA plot to the right, we notice the same effect in dimension 1 as in PC1, were samples are clearly separated by treatment. The similarities between timepoints are still noticeable, but the batch effect is not as evident. Additionally, we see that the `"auth"` samples for 72h are clustering away from the rest of the samples in similar manner.

We proceed now with the MDS plots. In the MDS plots the x- and y-axis are constructed in a way such that they preserve the high-dimensional distances between samples (based on the top 500 variable genes). The first step of the process is therefore, to calculate the high-dimensional distances. We provide Euclidean distances calculated earlier for the VST data and Poisson distances for the count data. **Note that regardless of the type of distance provided, the low-dimensional distances which MDS tries to generate so that they reflect high-dimensional distances are Euclidean.** Therefore, the interpretation of the second plot below requires caution.

```{r MDS}
# VST
mds = as.data.frame(colData(vsd))  %>%
  cbind(cmdscale(sampleDistMatrix))
mdsplot = ggplot(mds, aes(x = `1`, y = `2`, color = treatment, 
                          shape = batch)) +
  geom_point(size = 2, alpha = 0.65) +
  geom_text_repel(aes(label = timepoint), segment.linetype = 1,
                      segment.color = "black", segment.size = 0.6,
                  min.segment.length = 0,
                  size = 3.5, nudge_y = 0.03, nudge_x = 0.03, show.legend = FALSE) +
  ggtitle("Multidimensional scaling: VST") +
  theme_classic() +
  scale_color_manual(name = "treatment", values = c("orange", "purple4"),
                     labels = c("control", "HMBA"))+
  theme(plot.title = element_text(size = 12, face = "bold", vjust = 0.5, hjust = 0.5),
        axis.text = element_text(size = 8, hjust = 0.5, vjust = 0.5, 
                                 color = "black"),
        axis.title = element_text(size = 10, face = "bold"),
        axis.ticks = element_line(linewidth = 0.1),
        axis.line = element_line(linewidth = 0.3),
        legend.position = "right",
        legend.key.size = unit(6, units = "mm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(face = "bold", size = 8),
        legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "mm"),
        legend.spacing.y = unit(2, units = "mm"),
        legend.spacing.x = unit(2, units = "mm"),
        legend.background = element_blank())+
  labs(x = "MDS1", y = "MDS2") +
  guides(alpha = "none")

# Poisson
mdsPois = as.data.frame(colData(dds)) %>%
  cbind(cmdscale(samplePoisDistMatrix))
poisson_mdsplot = ggplot(mdsPois, aes(x = `1`, y = `2`, color = treatment, 
                                      shape = batch)) +
  geom_point(size = 2, alpha = 0.65) +
  geom_text_repel(aes(label = timepoint), segment.linetype = 1,
                      segment.color = "black", segment.size = 0.6,
                  min.segment.length = 0,
                  size = 3.5, nudge_y = 0.03, nudge_x = 0.03, show.legend = FALSE) +
  ggtitle("Multidimensional scaling: Poisson distance (counts)") +
  theme_classic() +
  scale_color_manual(name = "treatment", values = c("orange", "purple4"),
                     labels = c("control", "HMBA"))+
  theme(plot.title = element_text(size = 12, face = "bold", vjust = 0.5, hjust = 0.5),
        axis.text = element_text(size = 8, hjust = 0.5, vjust = 0.5, 
                                 color = "black"),
        axis.title = element_text(size = 10, face = "bold"),
        axis.ticks = element_line(linewidth = 0.1),
        axis.line = element_line(linewidth = 0.3),
        legend.position = "right",
        legend.key.size = unit(6, units = "mm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(face = "bold", size = 8),
        legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "mm"),
        legend.spacing.y = unit(2, units = "mm"),
        legend.spacing.x = unit(2, units = "mm"),
        legend.background = element_blank())+
  labs(x = "MDS1", y = "MDS2") +
  guides(alpha = "none")

fig = ggarrange(mdsplot +rremove("ylab"), poisson_mdsplot +rremove("ylab"), 
                ncol = 2, nrow = 1, 
                common.legend = TRUE, legend = "right",
                labels = c("A", "B"),
                align = "hv",
                font.label = list(size =4, color = "black", face = "bold"))
annotate_figure(fig, left = textGrob(bquote(bold("MDS2")), rot = 90, vjust = 1, gp = gpar(cex = 0.45)))
```
Here, we notice patterns similar to the ones from the PCA plots. In the left plot (VST data), we that MDS1 separates samples primarily based on treatment and MDS2 separates samples primarily based on batch (as in the VST PCA case). The MDS of the Poisson distances has a wildly different scale in both MDS1 and MDS2, but we still notice that MDS1 is dominated by treatment differences and MDS2 by batch.

Overall, from a first look in our data, it seems like sample differences are mainly affected by treatment and batch, while timepoint may have a milder but consistent effect.

# Differential Expression
## DESeq2 analysis
We are going to use a likelihood ratio test (LRT) as described in the [DESeq2 vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#time-series-experiments) and the [RNAseq workflow by Michael Love](http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#time-course-experiments). 

The LRT examines two models for the counts, a full model with a certain number of terms and a reduced model, in which some of the terms of the full model are removed. The test determines if the increased likelihood of the data using the extra terms in the full model is more than expected if those extra terms are truly zero.

The LRT is therefore useful for testing multiple terms at once, for example testing 3 or more levels of a factor at once, or all interactions between two variables. The LRT for count data is conceptually similar to an analysis of variance (ANOVA) calculation in linear regression, except that in the case of the Negative Binomial GLM, we use an analysis of deviance (ANODEV), where the deviance captures the difference in likelihood between a full and a reduced model. In our case, the reduced model does not contain the `treatment:timepoint` interaction term.

Genes with small $p$-values from this test are those which at one or more time points after time `24h` showed a treatment-specific effect. Note, therefore, that this will not give small $p$-values to genes that moved up or down over time in the same way in both treatments.

```{r dgea}
dds = DESeq(dds, test = "LRT", reduced = ~ batch + treatment + timepoint)
resultsNames(dds) # Names of the coefficients
```
The equation of our model can be written as:
$$\log(\mu_{ij}) = \beta_0 + \beta_1 \cdot \mathbb{1}{_\text{batch = 2023_exp}}(j) + \beta_2 \cdot \mathbb{1}{_\text{treatment = HMBA}}(j) + \beta_3 \cdot \mathbb{1}{_\text{timepoint = 48h}}(j) + \\ + \beta_4 \cdot \mathbb{1}{_\text{timepoint = 72h}}(j) + \gamma_1 \cdot \mathbb{1}{_\text{HMBA48h}} + \gamma_2 \cdot \mathbb{1}{_\text{HMBA72h}} + \log(s_j)$$
We are interested in the following comparisons:

  1. HMBA24h vs. Ctrl48h
  2. HMBA48h vs. Ctrl48h
  3. HMBA72h vs. Ctrl48h
  4. HMBA48h vs. HMBA24h
  5. HMBA72h vs. HMBA24h
  6. HMBA72h vs. HMBA48h
  
In order to perform these comparisons, we must specify our contrasts accordingly. Contrasts essentially compare coefficients or combinations of coefficients (for more complex designs such as this one here). In our model, the reference level for treatment is `control` and the reference level for timepoint is `24h`. 

For example, to compare HMBA24h vs. Ctrl48h, we must get the coefficients that describe the effect of HMBA at 24h and sum them, and then subtract the sum of the coefficients that describe the no-treatment samples at 48h. For HMBA at 24h we only need $\beta_2$ based on our equation, while for Ctrl at 48h, we only need $\beta_3$. Therefore, the contrast for HMBA24h vs. Ctrl48h is equal to $\beta_2 - \beta_3$. Similarly, for HMBA48h we need the overall effect of treatment, plus the effect of timepoint `48h`, plus the specific treatment-timepoint combo ($\beta_2 + \gamma_1 + \beta_3$) and therefore for the comparison HMBA48h vs. Ctrl48h, the contrast is $\beta_2 + \gamma_1 + \beta_3 - \beta_3 = \beta_2 + \gamma_1$. All other comparions follow a similar logic which is captured in the code below.

```{r contrasts}
contrasts = list(HMBA24h_vs_Control48h = list("treatment_HMBA_vs_control",
                 "timepoint_48h_vs_24h"),
                 HMBA48h_vs_Control48h = list(c("treatment_HMBA_vs_control", 
                                                "treatmentHMBA.timepoint48h")),
                 HMBA72h_vs_Control48h = list(c("treatment_HMBA_vs_control",
                                                "timepoint_72h_vs_24h",
                                                "treatmentHMBA.timepoint72h"),
                                              "timepoint_48h_vs_24h"),
                 HMBA48h_vs_HMBA24h = list(c("timepoint_48h_vs_24h",
                                              "treatmentHMBA.timepoint48h")),
                 HMBA72h_vs_HMBA24h = list(c("treatmentHMBA.timepoint72h",
                                             "timepoint_72h_vs_24h")),
                 HMBA72h_vs_HMBA48h = list(c("timepoint_48h_vs_24h",
                                             "treatmentHMBA.timepoint48h"),
                                           c("treatmentHMBA.timepoint72h",
                                             "timepoint_72h_vs_24h")))

# hmba24h vs control48h
res_hmba24h_vs_control48h = results(dds, contrast=contrasts[["HMBA24h_vs_Control48h"]], 
                                     listValues=c(1, -1), test = "Wald")

# hmba48h vs control48h
res_hmba48h_vs_control48h = results(dds, 
                                     contrast=contrasts[["HMBA48h_vs_Control48h"]], 
                                     listValues=c(1, -1), test = "Wald")

# hmba72h vs control48h
res_hmba72h_vs_control48h = results(dds, contrast=contrasts[["HMBA72h_vs_Control48h"]],
                                     listValues=c(1, -1), test = "Wald")

# HMBA48h vs HMBA24h
res_HMBA48h_vs_HMBA24h = results(dds, 
                                  contrast=contrasts[["HMBA48h_vs_HMBA24h"]],
                                  listValues=c(1, -1), test = "Wald")

# HMBA72h vs HMBA24h
res_HMBA72h_vs_HMBA24h = results(dds, 
                                  contrast=contrasts[["HMBA72h_vs_HMBA24h"]], 
                                  listValues=c(1, -1), test = "Wald")

# HMBA72h vs HMBA48h
res_HMBA72h_vs_HMBA48h = results(dds, contrast=contrasts[["HMBA72h_vs_HMBA48h"]],
                                  listValues=c(1, -1), test = "Wald")

results = list(res_hmba24h_vs_control48h, res_hmba48h_vs_control48h, 
               res_hmba72h_vs_control48h, res_HMBA48h_vs_HMBA24h,
               res_HMBA72h_vs_HMBA24h, res_HMBA72h_vs_HMBA48h)
names(results) = names(contrasts)
```

## Log-fold change shrinkage
A common problem in RNA-seq analysis is that genes with low counts may be associated with unexpectedly high log-fold change (LFC) statistics. `DESeq2` offers a couple of ways to shrink LFCs to yield more stable and robust results in a Bayesian way (by assessing the distribution of LFCs and shrinking values towards the central trend of LFCs). There are three ways to shrink LFCs in `DESeq2`: 

  - the adaptive t prior shrinkage estimator from the `apeglm` package, 
  - the adaptive shrinkage estimator from the `ashr` package and 
  - shrinkage based on a normal prior (see more [here](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#extended-section-on-shrinkage-estimators)).

Here we use `ashr` only, because `apeglm` does not work with specific contrasts and `normal` does not work with interaction designs.

```{r shrinkage}
ashr_shrunk_results = list()

for (i in 1:length(results)) {
  ashr_shrunk_results[[i]] = lfcShrink(dds, res = results[[i]], type = "ashr")
}
names(ashr_shrunk_results) = names(results)
```

We now provide code to export the results of differential expression analysis in MS Excel format using the `openxlsx` package.

```{r export dgea, eval = FALSE}
# Create a workbook and save data there to export later
dgea_wb = createWorkbook()
dgea_wb_shrunk = createWorkbook()

# Create data frames that will be exported into .xlsx files
for (i in 1:length(results)) {
  # Raw LFCs
  genes = rownames(results[[i]])
  res = results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj) %>%
    dplyr::select(Gene.Symbol, everything())
  
  addWorksheet(dgea_wb, names(results)[i])
  writeData(dgea_wb, names(results)[i], res)
  
  # Apply the styles
  addStyle(dgea_wb, names(results)[i], style = header_style, rows = 1, 
           cols = 1:ncol(res), gridExpand = TRUE) # first row only
  addStyle(dgea_wb, names(results)[i], style = body_style, rows = 2:nrow(res), 
           cols = 1:ncol(res), gridExpand = TRUE)
  
  # ASHR LFCs
  genes = rownames(ashr_shrunk_results[[i]])
  res_ashr = ashr_shrunk_results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj) %>%
    dplyr::select(Gene.Symbol, everything())
  
  addWorksheet(dgea_wb_shrunk, names(ashr_shrunk_results)[i])
  writeData(dgea_wb_shrunk, names(ashr_shrunk_results)[i], res_ashr)
  
  # Apply the styles
  addStyle(dgea_wb_shrunk, names(ashr_shrunk_results)[i], style = header_style, 
           rows = 1, cols = 1:ncol(res_ashr), gridExpand = TRUE) # first row only
  addStyle(dgea_wb_shrunk, names(ashr_shrunk_results)[i], style = body_style, 
           rows = 2:nrow(res_ashr), cols = 1:ncol(res_ashr), gridExpand = TRUE)
}

# Create DGEA folder
if (!dir.exists("DGEA")) {
  dir.create("DGEA")
}

saveWorkbook(dgea_wb, "DGEA/DGEA_raw.xlsx",
             overwrite = TRUE)
saveWorkbook(dgea_wb_shrunk, "DGEA/DGEA_ashr.xlsx",
             overwrite = TRUE)
rm(genes, res, res_ashr, i); gc()
```

# Post-DGEA plotting
## MA-plots
In the context of RNA-seq, an MA plot is a scatter plot where each point represents a gene:

  - The x-axis (A) displays the average of the log-transformed normalized counts across all samples or conditions for a given gene. This is a measure of the gene's overall expression level.
  - The y-axis (M) displays the LFC between two conditions for a given gene, indicating differential expression.
  
In essence, the MA plot allows you to see the relationship between the expression change of genes (LFC), which is plotted on the y-axis, and the average expression level of genes, which is plotted on the x-axis.

The main advantage of an MA plot is that it allows for easy visualization of the relationship between the magnitude of gene expression changes and the average expression level of genes. This is important because statistical tests for differential expression can be more sensitive for genes with higher overall expression levels, leading to a potential bias.

The following code is what we run in our main script to We produced MA plots for our all of our comparisons and they are available in the repository (in the `"DGEA/MA_plots"` folder).

Here we provide the MA plots for raw and shrunk LFCs for the HMBA24h vs. Ctrl48h comparison:
```{r MA_plot_example}
par(mfrow=c(1,2), mar=c(4,4,2,1))
  print(DESeq2::plotMA(results[[i]], main = "MA plot: raw LFCs",
                       xlab = substitute(paste(bold('Mean of normalized counts'))),
                       ylab = substitute(paste(bold('log fold change'))),
                       alpha = 0.05))
  print(DESeq2::plotMA(ashr_shrunk_results[[i]], main = "MA plot: ashr shrunk LFCs",
                       xlab = substitute(paste(bold('Mean of normalized counts'))),
                       ylab = substitute(paste(bold('log fold change'))),
                       alpha = 0.05))
```

## Volcano plots
Volcano plots are one of the most common ways to visualize results from Differential Expression Analysis. The x-axis in a volcano plot is usually centered at zero and displays the LFCs of the genes and the y-axis displays the $-\log_{10}(p_{adj_{BH}})$ (negative base-10 logarithm of the Benjamini-Hochberg adjusted $p$-value in our case).

Here we provide the volcano plots for all comparisons (also available in the `"DGEA/Volcano_plots"` folder): for raw LFCs first and then for the `ashr` LFCs. The top 20 genes (with the lowest $p_{adj_{BH}}$) are also labelled.
```{r raw_lfc_volcanos, fig.height = 10}
# Raw LFCs
for (i in 1:length(results)) {
  # Construct the data frame
  genes = rownames(results[[i]])
  res = results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  # create custom key-value pairs for stat. sig genes (p.adj < 0.05) and n.s genes
  keyvals.colour <- ifelse(
    res$log2FoldChange < -1 & res$padj < 0.05, 'royalblue',
    ifelse(res$log2FoldChange > 1 & res$padj < 0.05, 'red4',
           ifelse(abs(res$log2FoldChange) < 1 & res$padj < 0.05, 'pink',
                  'grey')))
  
  # keyvals.colour[is.na(keyvals.colour)] <- 'black'
  names(keyvals.colour)[keyvals.colour == 'royalblue'] <- 'Down-regulated'
  names(keyvals.colour)[keyvals.colour == 'red4'] <- 'Up-regulated'
  names(keyvals.colour)[keyvals.colour == 'pink'] <- '|log2FC| < 1'
  names(keyvals.colour)[keyvals.colour == 'grey'] <- 'p.adj > 0.05'
  
  res$aes = keyvals.colour
  
  volcano = EnhancedVolcano(res,
                            lab = res[, "Gene.Symbol"],
                            caption = NULL,
                            x = 'log2FoldChange',
                            y = 'padj',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            cutoffLineType = "dashed",
                            cutoffLineWidth = 0.5,
                            cutoffLineCol = "black",
                            colCustom = keyvals.colour,
                            colAlpha = 0.7,
                            ylab = bquote(bold(-log[10]("BH adj. p-value"))),
                            xlab = bquote(bold(log[2]("Fold Change"))),
                            pointSize = 1.5,
                            axisLabSize = 7,
                            subtitle = NULL,
                            labSize = 4,
                            selectLab = res[1:20, "Gene.Symbol"],
                            legendLabSize = 8,
                            legendIconSize = 4,
                            title = paste0("Volcano plot: contrast = ",
                            names(results)[i]),
                            labFace = "bold",
                            boxedLabels = TRUE,
                            drawConnectors = TRUE,
                            typeConnectors = "closed",
                            arrowheads = FALSE,
                            widthConnectors = 0.5,
                            max.overlaps = Inf) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.4),
          plot.title = element_text(face = "bold", size = 14, hjust = 0.5, vjust = 0.5),
          axis.title = element_text(face = "bold", size = 11),
          axis.text = element_text(size = 9),
          axis.line = element_line(colour = "black", linewidth = 0.7),
          axis.ticks = element_line(colour = "black", linewidth = 0.7),
          axis.ticks.length = unit(1, units = "mm"),
          legend.position = "right",
          legend.title = element_blank(),
          legend.margin = ggplot2::margin(1, 1, 1, 1, unit = "mm"),
          legend.spacing.y = unit(3, units = "mm"),
          legend.spacing.x = unit(3, units = "mm"),
          legend.background = element_blank(),
          legend.key.size = unit(1.5, units = "mm"),
          legend.text = element_text(size = 10),
          legend.box.background = element_rect(colour = "black"))
  print(volcano)
}
rm(genes, i, res, keyvals.colour, volcano)
```

```{r ashr_volcanos, fig.height = 10}
# ashr-shrunk LFCs
for (i in 1:length(ashr_shrunk_results)) {
  # Construct the data frame
  genes = rownames(ashr_shrunk_results[[i]])
  res = ashr_shrunk_results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  # create custom key-value pairs for stat. sig genes (p.adj < 0.05) and n.s genes
  keyvals.colour <- ifelse(
    res$log2FoldChange < -1 & res$padj < 0.05, 'royalblue',
    ifelse(res$log2FoldChange > 1 & res$padj < 0.05, 'red4',
           ifelse(abs(res$log2FoldChange) < 1 & res$padj < 0.05, 'pink',
                  'grey')))
  
  # keyvals.colour[is.na(keyvals.colour)] <- 'black'
  names(keyvals.colour)[keyvals.colour == 'royalblue'] <- 'Down-regulated'
  names(keyvals.colour)[keyvals.colour == 'red4'] <- 'Up-regulated'
  names(keyvals.colour)[keyvals.colour == 'pink'] <- '|log2FC| < 1'
  names(keyvals.colour)[keyvals.colour == 'grey'] <- 'p.adj > 0.05'
  
  res$aes = keyvals.colour
  
  volcano = EnhancedVolcano(res,
                            lab = res[, "Gene.Symbol"],
                            caption = NULL,
                            x = 'log2FoldChange',
                            y = 'padj',
                            pCutoff = 0.05,
                            FCcutoff = 1,
                            cutoffLineType = "dashed",
                            cutoffLineWidth = 0.5,
                            cutoffLineCol = "black",
                            colCustom = keyvals.colour,
                            colAlpha = 0.7,
                            ylab = bquote(bold(-log[10]("BH adj. p-value"))),
                            xlab = bquote(bold(log[2]("Fold Change"))),
                            pointSize = 1.5,
                            axisLabSize = 7,
                            subtitle = NULL,
                            labSize = 4,
                            selectLab = res[1:20, "Gene.Symbol"],
                            legendLabSize = 8,
                            legendIconSize = 4,
                            title = paste0("Volcano plot: contrast = ",
                                           names(results)[i]),
                            labFace = "bold",
                            boxedLabels = TRUE,
                            drawConnectors = TRUE,
                            typeConnectors = "closed",
                            arrowheads = FALSE,
                            widthConnectors = 0.5,
                            max.overlaps = Inf) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.4),
          plot.title = element_text(face = "bold", size = 14, hjust = 0.5, vjust = 0.5),
          axis.title = element_text(face = "bold", size = 11),
          axis.text = element_text(size = 9),
          axis.line = element_line(colour = "black", linewidth = 0.7),
          axis.ticks = element_line(colour = "black", linewidth = 0.7),
          axis.ticks.length = unit(1, units = "mm"),
          legend.position = "right",
          legend.title = element_blank(),
          legend.margin = ggplot2::margin(1, 1, 1, 1, unit = "mm"),
          legend.spacing.y = unit(3, units = "mm"),
          legend.spacing.x = unit(3, units = "mm"),
          legend.background = element_blank(),
          legend.key.size = unit(1.5, units = "mm"),
          legend.text = element_text(size = 10),
          legend.box.background = element_rect(colour = "black"))
  print(volcano)
}
rm(genes, i, res, keyvals.colour, volcano)
```

## Post-DGEA heatmaps
Finally, we produce heatmaps using our results from differential expression analysis. We define a parameter `n_degs` as the number of genes (rows) of the heatmaps. We set this to `20`. We then proceed with extract the top `n_degs` variable genes from the VST data and the top `n_degs` genes from the differential expression results (for each comparison and type of shrinkage). All the heatmaps are stored in the `"DGEA/Post_DGEA_heatmaps"` folder in the repository.

```{r n_degs}
# Number of features for plots
n_degs = 20
```

We generate a heatmap with clustered samples (VST data) for the top 20 variable genes:
```{r vst_post_dgea_heatmap, fig.height = 10}
# Top n_degs variable genes fromt the VST object
topVarGenes = head(order(rowVars(assay(vsd)), decreasing = TRUE), n_degs)
matvar  = assay(vsd)[topVarGenes, ]
matvar  = matvar - rowMeans(matvar)
colnames(matvar) = colnames(assay(vsd))
anno = as.data.frame(colData(vsd)[, c("batch", "treatment", "timepoint")])

# Create a custom palette with 20*14 = 280 color breaks (one for each tile)
colors2 = colorRampPalette(c(viridisLite::mako(280)[140:280],
                             rev(viridisLite::rocket(280)[140:280])))(280)

pheatmap(matvar, annotation_col = anno,
         col = colors2,
         fontsize = 8,
         cutree_col = 2,
         main = "Top 20 variable VST genes (centered) heatmap",
         annotation_colors = ann_colors,
         cluster_rows = FALSE)
```
One thing we instantly notice here, is that there's perfect clustering treatment-wise based on the top 20 variable genes in VST data.

Regarding the differential expression results, we created heatmaps which clustered samples and genes based on their similarity (Poisson distances between genes and Poisson distances between samples) and heatmaps where genes and samples are pre-ordered based on LFC and treatment respectively. We also force clustering to stop at two groups, to see if these correlate with treatment, batch or timepoint. Most of the times they correlate perfectly with treatment.

We first present the raw LFC heatmaps with clustered samples and clustered genes.
```{r heatmaps_raw_lfcs_clustered, fig.height = 10}
# Raw LFCs: clustered data
for (i in 1:length(results)) {
  # Construct the data frame
  genes = rownames(results[[i]])
  res = results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  topDEGs = res[which(res$padj < 0.05),]
  
  # Gene clustering with raw counts and Poisson distance
  topDEGs = topDEGs[1:n_degs, ]
  matcount = assay(dds)[topDEGs$Gene.Symbol,]
  DEGpoisd = PoissonDistance(t(counts(dds[topDEGs$Gene.Symbol,])))
  DEGsamplePoisDistMatrix = as.matrix(DEGpoisd$dd)
  rownames(DEGsamplePoisDistMatrix) = dds$cell_line
  colnames(DEGsamplePoisDistMatrix) = NULL
  
  # Same for genes
  DEGpoisd_genes = PoissonDistance(counts(dds[topDEGs$Gene.Symbol,]))
  DEGgenesPoisDistMatrix = as.matrix(DEGpoisd_genes$dd)
  rownames(DEGgenesPoisDistMatrix) = topDEGs$Gene.Symbol
  colnames(DEGgenesPoisDistMatrix) = NULL
  
  # Log-transformed matrix
  logmat = log2(matcount + 1)
  
  # Mean-centering
  logmat_znorm = (logmat - rowMeans(logmat))/sqrt(rowVars(logmat))
  
  # In the plot below the cluster distances are not affected by the log-transformation
  # of the matrix. It is only used to make a smoother color scale
  print(pheatmap(logmat_znorm, annotation_col = anno,
           breaks = seq(-4, 4, 8/length(colors2)),
           annotation_colors = ann_colors,
           clustering_distance_rows = DEGpoisd_genes$dd,
           clustering_distance_cols = DEGpoisd$dd,
           col = colors2,
           cutree_cols = 2,
           cutree_rows = 2,
           fontsize = 8,
           fontsize_row = 8,
           fontsize_col = 6,
           main = paste0(names(results)[i], 
                         " - DEGs vs. samples: Poisson heatmap, cluster cut = 2"),
           legend_breaks = c(-4, 4),
           legend_labels = c("lower expression", "higher expression")))
}
rm(genes, i, res, topDEGs, matcount, DEGgenesPoisDistMatrix,
   DEGpoisd, DEGpoisd_genes, DEGsamplePoisDistMatrix, logmat, logmat_znorm)
```

We first present the raw LFC heatmaps with ordered samples and ordered genes.

```{r heatmaps_raw_lfcs_ordered, fig.height = 10}
# Raw LFCs: ordered data
for (i in 1:length(results)) {
  # Construct the data frame
  genes = rownames(results[[i]])
  res = results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  topDEGs = res[which(res$padj < 0.05),]
  
  # Gene clustering with raw counts and Poisson distance
  topDEGs = topDEGs[1:n_degs, ]
  matcount = assay(dds)[topDEGs$Gene.Symbol,]
  DEGpoisd = PoissonDistance(t(counts(dds[topDEGs$Gene.Symbol,])))
  DEGsamplePoisDistMatrix = as.matrix(DEGpoisd$dd)
  rownames(DEGsamplePoisDistMatrix) = dds$cell_line
  colnames(DEGsamplePoisDistMatrix) = NULL
  
  # Same for genes
  DEGpoisd_genes = PoissonDistance(counts(dds[topDEGs$Gene.Symbol,]))
  DEGgenesPoisDistMatrix = as.matrix(DEGpoisd_genes$dd)
  rownames(DEGgenesPoisDistMatrix) = topDEGs$Gene.Symbol
  colnames(DEGgenesPoisDistMatrix) = NULL
  
  # Log-transformed matrix
  logmat = log2(matcount + 1)
  
  # The same matrix but order samples by metastatic status and genes by logFC (raw)
  genes_order = topDEGs %>% dplyr::filter(padj < 0.05) %>%
    arrange(log2FoldChange)
  gaps_row = length(which(genes_order$log2FoldChange < 0))
  genes_order = genes_order$Gene.Symbol
  
  sample_order = anno %>% dplyr::arrange(treatment)
  gaps_col = length(which(sample_order$treatment == "control"))
  sample_order = rownames(sample_order)
  
  matcount_ordered = matcount[genes_order, sample_order]
  
  # Mean-centering
  logmat_znorm = (logmat - rowMeans(logmat))/sqrt(rowVars(logmat))
  logmat_ordered_znorm = logmat_znorm[genes_order, sample_order]
  
  # In the plot below the cluster distances are not affected by the log-transformation
  # of the matrix. It is only used to make a smoother color scale
  print(pheatmap(logmat_ordered_znorm, annotation_col = anno,
           breaks = seq(-4, 4, 8/length(colors2)),
           annotation_colors = ann_colors,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           col = colors2,
           gaps_row = gaps_row,
           gaps_col = gaps_col,
           cutree_cols = 2,
           fontsize = 8,
           fontsize_row = 8,
           fontsize_col = 6,
           border_color = "grey60",
           main = paste0(names(results)[i], 
                         " - DEGs vs. samples heatmap (ordered)"),
           legend_breaks = c(-4, 4),
           legend_labels = c("lower expression", "higher expression")))
}
rm(genes, i, res, topDEGs, matcount, matcount_ordered, DEGgenesPoisDistMatrix,
   DEGpoisd, DEGpoisd_genes, DEGsamplePoisDistMatrix, logmat, logmat_ordered_znorm,
   logmat_znorm, sample_order, genes_order, gaps_col)
```

We now proceed with clustered ashr-shrunk data.
```{r heatmaps_ashr_lfcs_clustered, fig.height = 10}
# ashr LFCs: clustered data
for (i in 1:length(ashr_shrunk_results)) {
  # Construct the data frame
  genes = rownames(ashr_shrunk_results[[i]])
  res = ashr_shrunk_results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  topDEGs = res[which(res$padj < 0.05),]
  
  # Gene clustering with raw counts and Poisson distance
  topDEGs = topDEGs[1:n_degs, ]
  matcount = assay(dds)[topDEGs$Gene.Symbol,]
  DEGpoisd = PoissonDistance(t(counts(dds[topDEGs$Gene.Symbol,])))
  DEGsamplePoisDistMatrix = as.matrix(DEGpoisd$dd)
  rownames(DEGsamplePoisDistMatrix) = dds$cell_line
  colnames(DEGsamplePoisDistMatrix) = NULL
  
  # Same for genes
  DEGpoisd_genes = PoissonDistance(counts(dds[topDEGs$Gene.Symbol,]))
  DEGgenesPoisDistMatrix = as.matrix(DEGpoisd_genes$dd)
  rownames(DEGgenesPoisDistMatrix) = topDEGs$Gene.Symbol
  colnames(DEGgenesPoisDistMatrix) = NULL
  
  # Log-transformed matrix
  logmat = log2(matcount + 1)
  
  # Mean-centering
  logmat_znorm = (logmat - rowMeans(logmat))/sqrt(rowVars(logmat))
  
  # In the plot below the cluster distances are not affected by the log-transformation
  # of the matrix. It is only used to make a smoother color scale
  print(pheatmap(logmat_znorm, annotation_col = anno,
           breaks = seq(-4, 4, 8/length(colors2)),
           annotation_colors = ann_colors,
           clustering_distance_rows = DEGpoisd_genes$dd,
           clustering_distance_cols = DEGpoisd$dd,
           col = colors2,
           cutree_cols = 2,
           cutree_rows = 2,
           fontsize = 8,
           fontsize_row = 8,
           fontsize_col = 6,
           main = paste0(names(ashr_shrunk_results)[i], 
                         " - DEGs vs. samples: Poisson heatmap, cluster cut = 2"),
           legend_breaks = c(-4, 4),
           legend_labels = c("lower expression", "higher expression")))
}
rm(genes, i, res, topDEGs, matcount, DEGgenesPoisDistMatrix,
   DEGpoisd, DEGpoisd_genes, DEGsamplePoisDistMatrix, logmat, logmat_znorm)
```

...and finally with ordered ashr-shrunk data.
```{r heatmaps_ashr_lfcs_ordered, fig.height = 10}
# ashr LFCs: ordered data
for (i in 1:length(ashr_shrunk_results)) {
  # Construct the data frame
  genes = rownames(ashr_shrunk_results[[i]])
  res = ashr_shrunk_results[[i]]@listData %>%
    as.data.frame() %>% 
    mutate(Gene.Symbol = genes) %>%
    dplyr::arrange(padj)
  
  topDEGs = res[which(res$padj < 0.05),]
  
  # Gene clustering with raw counts and Poisson distance
  topDEGs = topDEGs[1:n_degs, ]
  matcount = assay(dds)[topDEGs$Gene.Symbol,]
  DEGpoisd = PoissonDistance(t(counts(dds[topDEGs$Gene.Symbol,])))
  DEGsamplePoisDistMatrix = as.matrix(DEGpoisd$dd)
  rownames(DEGsamplePoisDistMatrix) = dds$cell_line
  colnames(DEGsamplePoisDistMatrix) = NULL
  
  # Same for genes
  DEGpoisd_genes = PoissonDistance(counts(dds[topDEGs$Gene.Symbol,]))
  DEGgenesPoisDistMatrix = as.matrix(DEGpoisd_genes$dd)
  rownames(DEGgenesPoisDistMatrix) = topDEGs$Gene.Symbol
  colnames(DEGgenesPoisDistMatrix) = NULL
  
  # Log-transformed matrix
  logmat = log2(matcount + 1)
  
  # The same matrix but order samples by metastatic status and genes by logFC (raw)
  genes_order = topDEGs %>% dplyr::filter(padj < 0.05) %>%
    arrange(log2FoldChange)
  gaps_row = length(which(genes_order$log2FoldChange < 0))
  genes_order = genes_order$Gene.Symbol
  
  sample_order = anno %>% dplyr::arrange(treatment)
  gaps_col = length(which(sample_order$treatment == "control"))
  sample_order = rownames(sample_order)
  
  matcount_ordered = matcount[genes_order, sample_order]
  
  # Mean-centering
  logmat_znorm = (logmat - rowMeans(logmat))/sqrt(rowVars(logmat))
  logmat_ordered_znorm = logmat_znorm[genes_order, sample_order]
  
  # In the plot below the cluster distances are not affected by the log-transformation
  # of the matrix. It is only used to make a smoother color scale

  # Ordered genes and samples
  print(pheatmap(logmat_ordered_znorm, annotation_col = anno,
           breaks = seq(-4, 4, 8/length(colors2)),
           annotation_colors = ann_colors,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           col = colors2,
           gaps_row = gaps_row,
           gaps_col = gaps_col,
           cutree_cols = 2,
           fontsize = 8,
           fontsize_row = 8,
           fontsize_col = 6,
           border_color = "grey60",
           main = paste0(names(ashr_shrunk_results)[i], 
                         " - DEGs vs. samples heatmap (ordered)"),
           legend_breaks = c(-4, 4),
           legend_labels = c("lower expression", "higher expression")))
}

rm(genes, i, res, topDEGs, matcount, matcount_ordered, DEGgenesPoisDistMatrix,
   DEGpoisd, DEGpoisd_genes, DEGsamplePoisDistMatrix, logmat, logmat_ordered_znorm,
   logmat_znorm, sample_order, genes_order, gaps_col)
```

# Session info
```{r session_info}
sessionInfo()
```